---
title: "hiPSCi cell line evaluation (hNIL) - 03 Normalisation"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=4, fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)
```

```{r, warning=FALSE}
# Load required packages
library(DropletUtils)
library(Matrix)
library(scater) 
library(scran)
library(BiocSingular)
library(BiocParallel)
library(pheatmap)
library(plotly)
library(batchelor)
library(viridis)
library(reshape2)
library(SingleCellExperiment)
library(cowplot)
library(dplyr)
ncores = 3
mcparam = MulticoreParam(workers = ncores)
register(mcparam)
```

For this experiment, we have 4 samples which have been multiplexed by genotype. We have two different pooling strategy used:

 1. Pool 1 and 2 conditions: The 8 ipsc lines were differentiated in separate wells (n=2, total of 16 wells) from day 1 to 18. On day 18 the cells were papain-dissociated and pooled the eight lines into one single tube (n=2) to proceed with 10x genomics experiment.
 2. Combined 1 and 2 conditions: The 8 ipsc lines were differentiated in separate wells (n=2, total of 16 wells) from day 1 to 3.  On day 3 the 8 lines were pooled into one well (n=2) and continue differentiation until day 18.  On day 17 the cells were papain-dissociated to proceed with 10x genomics experiment.

- Sample Pool1: hNIL pool cell line 1-8 replicate 1
- Sample Pool2: hNIL pool cell line 1-8 replicate 2
- Sample Combined1: hNIL combined cell line 1-8 replicate 1
- Sample Combined2: hNIL combined cell line 1-8 replicate 2

# Normalisation

```{r}
merged_sce_object = readRDS("qc_cell.rds")
```

## Merged normalisation

We will now do normalisation between samples.

```{r, fig.height=3}
set.seed(1000)
clusters = quickCluster(merged_sce_object, use.ranks=FALSE, BSPARAM=IrlbaParam(), BPPARAM=mcparam)
table(clusters)
table(merged_sce_object$Sample, clusters)
ggplot(melt(table(merged_sce_object$Sample, clusters)), aes(x=clusters, y=value, fill=Var1)) + geom_bar(stat="identity") + theme_cowplot()
```

```{r}
merged_sce_object = computeSumFactors(merged_sce_object, min.mean=0.1, cluster=clusters, BPPARAM=mcparam)
```

Let's visualise the size factors.

```{r, fig.height=4}
merged_sce_object_mixed = merged_sce_object[, sample(ncol(merged_sce_object))]
plotColData(merged_sce_object_mixed, x="sum", y=I(sizeFactors(merged_sce_object_mixed)), colour_by = "Sample") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Number of UMIs", y = "Size Factor") 
```

```{r, fig.width=9, fig.height=8}
plotColData(merged_sce_object_mixed, x="sum", y=I(sizeFactors(merged_sce_object_mixed)), colour_by = "Sample", other_fields = "Sample") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Number of UMIs", y = "Size Factor") +
  facet_wrap(~Sample, ncol=2)
```

Let's do the normalisation now.

```{r}
merged_sce_object_norm = logNormCounts(merged_sce_object)
```

Let's have a quick look at the distribution of the normalised counts.

```{r}
merged_sce_object_norm_qc = addPerCellQC(merged_sce_object_norm, exprs_values = "logcounts")
colData(merged_sce_object_norm_qc) = colData(merged_sce_object_norm_qc)[, c(1:3, (length(colData(merged_sce_object_norm))+1):length(colData(merged_sce_object_norm_qc)))]
```

```{r}
plotColData(merged_sce_object_norm_qc, x = "Sample", y = "sum") +
  labs(y = "normalised UMI count")
```

We see a lower distribution in the normalised count of the pooled sample compared to the combined sample. However, the difference is not too big (<10%), so it should not effect downstream analysis too much. 

TODO: double check with John

```{r}
saveRDS(merged_sce_object_norm, "norm_cell.rds")
```
