---
title: "hiPSCi cell line evaluation (hypothalamic) - 03 Normalisation"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=4, fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)
```

```{r, warning=FALSE}
# Load required packages
library(DropletUtils)
library(Matrix)
library(scater) 
library(scran)
library(BiocSingular)
library(BiocParallel)
library(pheatmap)
library(plotly)
library(batchelor)
library(viridis)
library(reshape2)
library(SingleCellExperiment)
library(cowplot)
library(dplyr)
ncores = 3
mcparam = MulticoreParam(workers = ncores)
register(mcparam)
```

For this experiment, we have 2 samples which have been multiplexed by both genotype and replicate (using multiseq sample labelling). In the cortical differentiation, all the cell lines were pooled together before differentiation. Furthermore, we have 6 replicates overall from 3 different operator with 2 technical replicates each.

- Sample SIGAE1: Cortical differentiation cell line 1-8 user 1-3 technical replicate 1 
- Sample SIGAF1: Cortical differentiation cell line 1-8 user 1-3 technical replicate 2

# Normalisation

```{r}
merged_sce_object = readRDS("qc_cell_multiseq.rds")
```

## Merged normalisation

We will now do normalisation between samples.

```{r, fig.height=3}
set.seed(1000)
clusters = quickCluster(merged_sce_object, use.ranks=FALSE, BSPARAM=IrlbaParam(), BPPARAM=mcparam)
table(clusters)
table(merged_sce_object$Sample, clusters)
ggplot(melt(table(merged_sce_object$Sample, clusters)), aes(x=clusters, y=value, fill=Var1)) + geom_bar(stat="identity") + theme_cowplot()
```

```{r}
merged_sce_object = computeSumFactors(merged_sce_object, min.mean=0.1, cluster=clusters, BPPARAM=mcparam)
```

Let's visualise the size factors.

```{r, fig.height=4}
merged_sce_object_mixed = merged_sce_object[, sample(ncol(merged_sce_object))]
plotColData(merged_sce_object_mixed, x="sum", y=I(sizeFactors(merged_sce_object_mixed)), colour_by = "Sample") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Number of UMIs", y = "Size Factor") 
```
We see a slight shift between the two samples, with the SIGAF1 sample having higher distribution compared to SIGAE1 sample.

```{r, fig.width=9}
plotColData(merged_sce_object_mixed, x="sum", y=I(sizeFactors(merged_sce_object_mixed)), colour_by = "Sample", other_fields = "Sample") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Number of UMIs", y = "Size Factor") +
  facet_wrap(~Sample)
```

Let's do the normalisation now.

```{r}
merged_sce_object_norm = logNormCounts(merged_sce_object)
```

Let's have a quick look at the distribution of the normalised counts.

```{r}
merged_sce_object_norm_qc = addPerCellQC(merged_sce_object_norm, exprs_values = "logcounts")
colData(merged_sce_object_norm_qc) = colData(merged_sce_object_norm_qc)[, c(1:3, (length(colData(merged_sce_object_norm))+1):length(colData(merged_sce_object_norm_qc)))]
```

```{r}
plotColData(merged_sce_object_norm_qc, x = "Sample", y = "sum") +
  labs(y = "normalised UMI count")
```
The result of the normalisation looks fine, though there is a slight difference between the distribution of the two samples.

```{r}
saveRDS(merged_sce_object_norm, "norm_cell.rds")
```
