---
title: "hiPSCi cell line evaluation (cortical) - 05 Batch correction"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=4, fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)
```

```{r, warning=FALSE}
# Load required packages
library(DropletUtils)
library(Matrix)
library(scater) 
library(scran)
library(BiocSingular)
library(BiocParallel)
library(pheatmap)
library(plotly)
library(batchelor)
library(viridis)
library(reshape2)
library(SingleCellExperiment)
library(cowplot)
library(dplyr)
theme_set(theme_cowplot())
ncores = 3
mcparam = MulticoreParam(workers = ncores)
register(mcparam)
```

For this experiment, we have 2 samples which have been multiplexed by both genotype and replicate (using multiseq sample labelling). In the cortical differentiation, all the cell lines were pooled together before differentiation. Furthermore, we have 6 replicates overall from 3 different operator with 2 technical replicates each.

- Sample SIGAE1: Cortical differentiation cell line 1-8 user 1-3 technical replicate 1 
- Sample SIGAF1: Cortical differentiation cell line 1-8 user 1-3 technical replicate 2

# Batch correction

```{r}
merged_sce_object = readRDS("dc_cell.rds")

# To properly perform batch correction, we need to remove doublets first
merged_sce_object = merged_sce_object[, colData(merged_sce_object)$final_doublet_call == "Singlet"]
# After the first round of clustering and analysis, we detect that 1 user replicates (Bar 9 and 10) behave very differently compared to other replicates. As such, we will remove this user replicate.
merged_sce_object = merged_sce_object[, !(colData(merged_sce_object)$multiseq_assignment %in% c("Bar9", "Bar10"))]
```

For the modeling of gene-variance relationship, we will again remove MT and Ribosomal genes due to the issue we see during doublet detection.

```{r, fig.height=3}
set.seed(1000)
non_mito_ribo_genes = rownames(merged_sce_object)[grep(pattern = "^(MT-|RP[LS]|MALAT1)", rownames(merged_sce_object), invert = TRUE)]
dec_combined = modelGeneVar(merged_sce_object, subset.row=non_mito_ribo_genes, block=merged_sce_object$Sample)
hvg_combined = getTopHVGs(dec_combined)
length(hvg_combined)
```

```{r, fig.height=2}
trend_plotlist = lapply(1:length(dec_combined$per.block), function (i){
  sample = names(dec_combined$per.block)[[i]]
  fit = metadata(dec_combined$per.block[[i]])
ggplot(data = data.frame(mean = fit$mean, var = fit$var, name=ifelse(fit$mean > 4 | fit$var > 2, names(fit$mean), NA)),
                mapping = aes(x = mean, y = var, label=name)) +
  geom_point() +
  ggtitle(sample) +
  labs(x = "Mean log-expression", y = "Variance of log-expression") +
  stat_function(fun=fit$trend, colour="dodgerblue", size=1)
})

plot_grid(plotlist = trend_plotlist, ncol=2)
```

```{r}
set.seed(1000)
sce_object_mnn = correctExperiments(merged_sce_object, 
                                    batch = merged_sce_object$Sample,
                                    subset.row = hvg_combined,
                                    PARAM=FastMnnParam(
                                      merge.order=list("Sample_SIGAE1", "Sample_SIGAF1")
                                    ))
```

Let's run UMAP for visualisation.

```{r}
set.seed(1000)
sce_object_mnn = runUMAP(sce_object_mnn, ncomponents=2, dimred="corrected", spread = 1, min_dist = 0.4)
```

Let's also calculate the UMAP for the unmerged dataset so we can do comparison.

```{r}
set.seed(1000)
merged_sce_object = runPCA(merged_sce_object, subset_row=hvg_combined)
merged_sce_object = runUMAP(merged_sce_object, ncomponents=2, dimred="PCA", spread = 1, min_dist = 0.4)
```

```{r, fig.width=10, fig.height=5}
plot_grid(plotUMAP(merged_sce_object, colour_by="Sample") + ggtitle("Original"), 
          plotUMAP(sce_object_mnn, colour_by="Sample") + ggtitle("Final"))
```

```{r}
plotUMAP(merged_sce_object, colour_by="Sample", other_fields="Sample") + ggtitle("Original") + facet_wrap(~Sample) + theme(legend.position = "none")
plotUMAP(sce_object_mnn, colour_by="Sample", other_fields="Sample") + ggtitle("Final") + facet_wrap(~Sample) + theme(legend.position = "none")
```

```{r}
saveRDS(sce_object_mnn, "all_corrections.rds")
```


