---
title: "hiPSCi cell line evaluation (cortical) - 08 Defining cluster"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=4, fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)
```

```{r, warning=FALSE}
# Load required packages
library(DropletUtils)
library(Matrix)
library(scater) 
library(scran)
library(BiocSingular)
library(BiocParallel)
library(pheatmap)
library(plotly)
library(batchelor)
library(viridis)
library(reshape2)
library(SingleCellExperiment)
library(cowplot)
library(igraph)
library(dplyr)
theme_set(theme_cowplot())
ncores = 3
mcparam = MulticoreParam(workers = ncores)
register(mcparam)
```

```{r}
color_palette = c("#c4eadd",
                  "#d098d7",
                  "#dce496",
                  "#bab5ff",
                  "#f9d08b",
                  "#5bb4ef",
                  "#e1ad71",
                  "#16c4e8",
                  "#ffa3c6",
                  "#4ec9b1",
                  "#de9a98",
                  "#22c0ba",
                  "#ffbfb2",
                  "#62be93",
                  "#ffcfe0",
                  "#c0ffdb",
                  "#a5b17a",
                  "#abebff",
                  "#e0ffd0",
                  "#9dc4b8",
                  "#affff5")
```

For this experiment, we have 2 samples which have been multiplexed by both genotype and replicate (using multiseq sample labelling). In the cortical differentiation, all the cell lines were pooled together before differentiation. Furthermore, we have 6 replicates overall from 3 different operator with 2 technical replicates each.

- Sample SIGAE1: Cortical differentiation cell line 1-8 user 1-3 technical replicate 1 
- Sample SIGAF1: Cortical differentiation cell line 1-8 user 1-3 technical replicate 2

# Defining cluster

```{r}
sce_object_mnn = readRDS("all_corrections.rds")
```

We will now define the clusters to be used for further analysis. The clustering approach we will be using is essentially the same as the one we have previously performed for previous analysis. We first construct shared nearest neighbour graph and then identify the community using Louvain algorithm.

```{r, fig.height=3}
set.seed(1000)
snn_graph = buildSNNGraph(sce_object_mnn, use.dimred="corrected")
clusters = cluster_louvain(snn_graph)$membership
table(clusters)
```

```{r}
ggplot(as.data.frame(table(clusters)), aes(x=clusters, y=Freq)) +
    geom_bar(stat="identity") +
    labs(x = "Cluster number", y = "Number of cells", title = paste("Louvain cluster"))
```

```{r, fig.width=10, fig.height=5}
sce_object_mnn$cluster = as.factor(clusters)
plot_grid(plotUMAP(sce_object_mnn, colour_by="cluster", text_by="cluster") + scale_fill_manual(values = color_palette) + scale_color_manual(values = color_palette), plotUMAP(sce_object_mnn, colour_by="Sample"))
```

Let's check the contribution of each sample for the clustering.

```{r, fig.width=10}
plotUMAP(sce_object_mnn, colour_by="cluster", other_fields = "Sample") + facet_wrap(~Sample)
```

```{r}
table(sample=sce_object_mnn$Sample, cluster=sce_object_mnn$cluster)
```

Let's visualise this to make it easier to interpret.

```{r, fig.width=6, fig.height=4}
contributions = as.data.frame.table(table(sample=sce_object_mnn$Sample, cluster=sce_object_mnn$cluster))
colnames(contributions) = c("Sample", "Cluster", "Count")
  
ggplot(data=contributions, mapping = aes(x=Cluster, y=Count, fill=Sample)) +
geom_bar(stat="identity") + 
scale_x_discrete() +
labs(x=NULL)+
geom_text(aes(label=Count), position=position_stack(vjust=0.5), colour="white") +
labs(fill="") +
scale_fill_brewer(palette = "Set2", type = "qual")
```

```{r}
saveRDS(sce_object_mnn, "clustered_cell.rds")
```


# Confounding factors

We want to check if the clusters could be generated due to technical differences between the cells.

## Library size

```{r}
plotColData(sce_object_mnn, x="cluster", y="total", colour_by = "Sample") +
  labs(x = "Cluster", y = "#UMIs") 
```

Let's have a quick look at the contribution of different batches to see if the UMI size of some cluster is specific to certain batch.

```{r, fig.width=10}
plotColData(sce_object_mnn, x="cluster", y="total", colour_by = "Sample", other_fields = "Sample") +
  labs(x = "Cluster", y = "#UMIs") + 
  facet_wrap(~Sample)
```

Let's look based on cell line assignment.

```{r}
plotColData(sce_object_mnn, x="cluster", y="total", colour_by = "vireo_assignment") +
  labs(x = "Cluster", y = "#UMIs") + scale_fill_manual(values=pals::tol(length(unique(sce_object_mnn$vireo_assignment))))
```

```{r, fig.width=10}
plotColData(sce_object_mnn, x="cluster", y="total", colour_by = "vireo_assignment", other_fields = "vireo_assignment") +
  labs(x = "Cluster", y = "#UMIs") + 
  facet_wrap(~vireo_assignment, ncol=4) + scale_fill_manual(values=pals::tol(length(unique(sce_object_mnn$vireo_assignment))))
```

Let's also look based on user replicate (MULTIseq barcode).

```{r}
multiseq_barcodes = c("Bar1", "Bar2", "Bar3", "Bar4", "Bar5", "Bar6", "Bar7", "Bar8", "Bar9", "Bar10", "Bar11", "Bar12")
multiseq_barcodes = intersect(multiseq_barcodes, unique(sce_object_mnn$multiseq_assignment))
sce_object_mnn$multiseq_assignment = factor(sce_object_mnn$multiseq_assignment, levels=multiseq_barcodes)

plotColData(sce_object_mnn, x="cluster", y="total", colour_by = "multiseq_assignment") +
  labs(x = "Cluster", y = "#UMIs") + scale_fill_manual(values=pals::okabe(length(unique(sce_object_mnn$multiseq_assignment))))
```

```{r, fig.width=7.5}
plotColData(sce_object_mnn, x="cluster", y="total", colour_by = "multiseq_assignment", other_fields = "multiseq_assignment") +
  labs(x = "Cluster", y = "#UMIs") + 
  facet_wrap(~multiseq_assignment, ncol=3) + scale_fill_manual(values=pals::okabe(length(unique(sce_object_mnn$multiseq_assignment))))
```


## Cell line

Let's also see the contribution from the different cell lines.

```{r, fig.width=10, fig.height=5}
sce_object_mnn$cluster = as.factor(clusters)
plot_grid(plotUMAP(sce_object_mnn, colour_by="cluster", text_by="cluster") + scale_fill_manual(values = color_palette) + scale_color_manual(values = color_palette), plotUMAP(sce_object_mnn, colour_by="vireo_assignment") + scale_fill_manual(values=pals::tol(length(unique(sce_object_mnn$vireo_assignment)))))
```

```{r, fig.width=10}
plotUMAP(sce_object_mnn, colour_by="cluster", other_fields = "vireo_assignment") + scale_fill_manual(values = color_palette) + facet_wrap(~vireo_assignment, ncol=4)
```

```{r}
table(sample=sce_object_mnn$vireo_assignment, cluster=sce_object_mnn$cluster)
```

```{r, fig.width=4, fig.height=4}
sample_contributions = as.data.frame.table(table(sample=sce_object_mnn$Sample, cell_line=sce_object_mnn$vireo_assignment))
colnames(sample_contributions) = c("Sample", "Cell line", "Count")
  
ggplot(data=sample_contributions, mapping = aes(x=Sample, y=Count, fill=`Cell line`)) +
geom_bar(stat="identity") + 
scale_x_discrete() +
labs(x=NULL)+
geom_text(aes(label=Count), position=position_stack(vjust=0.5), colour="white") +
labs(fill="") +
scale_fill_manual(values=pals::tol(length(unique(sce_object_mnn$vireo_assignment))))
```

```{r, fig.width=6, fig.height=4}
contributions = as.data.frame.table(table(sample=sce_object_mnn$vireo_assignment, cluster=sce_object_mnn$cluster))
colnames(contributions) = c("Cell line", "Cluster", "Count")
  
ggplot(data=contributions, mapping = aes(x=Cluster, y=Count, fill=`Cell line`)) +
geom_bar(stat="identity") + 
scale_x_discrete() +
labs(x=NULL)+
geom_text(aes(label=Count), position=position_stack(vjust=0.5), colour="white") +
labs(fill="") +
scale_fill_manual(values=pals::tol(length(unique(sce_object_mnn$vireo_assignment))))
```

## User replicate (MULTIseq barcode)

Let's also see the contribution from the different user replicate

```{r, fig.width=10, fig.height=5}
multiseq_barcodes = c("Bar1", "Bar2", "Bar3", "Bar4", "Bar5", "Bar6", "Bar7", "Bar8", "Bar9", "Bar10", "Bar11", "Bar12")
multiseq_barcodes = intersect(multiseq_barcodes, unique(sce_object_mnn$multiseq_assignment))
sce_object_mnn$multiseq_assignment = factor(sce_object_mnn$multiseq_assignment, levels=multiseq_barcodes)
sce_object_mnn$cluster = as.factor(clusters)
plot_grid(plotUMAP(sce_object_mnn, colour_by="cluster", text_by="cluster") + scale_fill_manual(values = color_palette) + scale_color_manual(values = color_palette), plotUMAP(sce_object_mnn, colour_by="multiseq_assignment") + scale_fill_manual(values=pals::okabe(length(unique(sce_object_mnn$multiseq_assignment)))))
```

```{r, fig.width=7.5}
plotUMAP(sce_object_mnn, colour_by="cluster", other_fields = "multiseq_assignment") + scale_fill_manual(values = color_palette) + facet_wrap(~multiseq_assignment, ncol=3)
```

```{r}
table(sample=sce_object_mnn$multiseq_assignment, cluster=sce_object_mnn$cluster)
```

```{r, fig.width=4, fig.height=4}
sample_contributions = as.data.frame.table(table(sample=sce_object_mnn$Sample, cell_line=sce_object_mnn$multiseq_assignment))
colnames(sample_contributions) = c("Sample", "Cell line", "Count")
  
ggplot(data=sample_contributions, mapping = aes(x=Sample, y=Count, fill=`Cell line`)) +
geom_bar(stat="identity") + 
scale_x_discrete() +
labs(x=NULL)+
geom_text(aes(label=Count), position=position_stack(vjust=0.5), colour="white") +
labs(fill="") +
scale_fill_manual(values=pals::okabe(length(unique(sce_object_mnn$multiseq_assignment))))
```

```{r, fig.width=6, fig.height=4}
contributions = as.data.frame.table(table(sample=sce_object_mnn$multiseq_assignment, cluster=sce_object_mnn$cluster))
colnames(contributions) = c("MULTIseq Barcode", "Cluster", "Count")
  
ggplot(data=contributions, mapping = aes(x=Cluster, y=Count, fill=`MULTIseq Barcode`)) +
geom_bar(stat="identity") + 
scale_x_discrete() +
labs(x=NULL)+
geom_text(aes(label=Count), position=position_stack(vjust=0.5), colour="white") +
labs(fill="") +
scale_fill_manual(values=pals::okabe(length(unique(sce_object_mnn$multiseq_assignment))))
```

## Mitochondrial gene fraction

```{r}
plotColData(sce_object_mnn, x="cluster", y="subsets_Mito_percent", colour_by = "vireo_assignment", other_fields = "vireo_assignment") +
  labs(x = "Cluster", y = "Mitochrondrial UMI fraction") + scale_fill_manual(values=pals::tol(length(unique(sce_object_mnn$vireo_assignment))))
```

## Ribosomal gene fraction

```{r}
plotColData(sce_object_mnn, x="cluster", y="subsets_Ribo_percent", colour_by = "vireo_assignment", other_fields = "vireo_assignment") +
  labs(x = "Cluster", y = "Ribosomal UMI fraction") + scale_fill_manual(values=pals::tol(length(unique(sce_object_mnn$vireo_assignment))))
```

