---
title: "hiPSCi cell line evaluation (all differentiations) - Differentiation annotation and marker expression"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=4, fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)
```

```{r, warning=FALSE}
# Load required packages
library(DropletUtils)
library(Matrix)
library(scater) 
library(scran)
library(BiocSingular)
library(BiocParallel)
library(pheatmap)
library(plotly)
library(batchelor)
library(viridis)
library(reshape2)
library(SingleCellExperiment)
library(cowplot)
library(igraph)
library(dplyr)
#library(schex)
library(wesanderson)
library(ComplexHeatmap)
library(circlize)
library(pals)
library(ggrastr)
library(ggedit)
theme_set(theme_cowplot())
ncores = 3
mcparam = MulticoreParam(workers = ncores)
register(mcparam)
```

```{r}
color_palette = list(
  "hypothalamus"=c("#54c2af", "#c24dac", "#52bc59", "#9c5bce", "#98b946", "#626cd4", "#cba94a", "#6471b6", "#5b8436", "#d34680", "#3f8f64", "#ce473e", "#5ea5db", "#cf7941", "#c584c4", "#887031", "#ba6271"),
  "cortical"=c("#96e97c", "#074d65", "#23dbe1", "#1b4dab", "#a6b6f9", "#762aac", "#997cfb", "#285d28", "#d3d6c1", "#6e390d", "#65a10e", "#d60724", "#e3d769"),
  "hNGN2"=c("#c3ae81", "#7e8cdc", "#a3d37d", "#cd8fd5", "#57ad6a", "#de7b70", "#78d6af", "#d78ea8", "#9aa850", "#85aad6", "#cf9554","#63c8d7", "#62a38d"),
  "hNIL"=c("#75859d", "#5c0d47", "#539322", "#4a1999", "#4081ec", "#1b511d", "#f90da0", "#173d61", "#c7628e", "#2f937a", "#6b2402", "#9c6ae4", "#473a0a")
)

cell_Line_palette = c("#999999", "#5566AA", "#558866", "#AADDCC", "#DDCC66","#99bb55", "#BB0011")
```

# Marker gene expression

We want to evaluate the performance of each cell line in terms of differentiation efficiency to target cell type across all differentiations.

For each differentiations, we have annotated each clusters with their cell type and class of cell type (target neuron, neuron or progenitor/other).

## Data loading

Let's first load the annotation

```{r}
cluster_annotations = read.csv("cluster_annotation_16052021.csv", stringsAsFactors = F)
cluster_annotations$ClusterID = as.integer(gsub("[a-zA-Z]+", "", cluster_annotations$Cluster))
```

Let's load all the differentiations and assign the cell identity to each differentiations.

```{r}
differentiations = c("hypothalamus", "cortical", "hNGN2", "hNIL")
differentiations_sce_objects = lapply(differentiations, function(d){
  merged_sce_object = readRDS(paste0(d, "/dc_cell.rds"))
  sce_object_mnn = readRDS(paste0(d, "/clustered_cell.rds"))
  
  # Subset the merged_sce_object to include only cells found in sce_object_mnn
  merged_sce_object = merged_sce_object[, paste0(merged_sce_object$Sample, ".", merged_sce_object$Barcode) %in% paste0(sce_object_mnn$Sample, ".", sce_object_mnn$Barcode)]
  
  # Rearrange the sce_object_mnn to match the ordering of merged_sce_object
  colnames(sce_object_mnn) = paste0(sce_object_mnn$Sample, ".", sce_object_mnn$Barcode)
  sce_object_mnn = sce_object_mnn[, paste0(merged_sce_object$Sample, ".", merged_sce_object$Barcode)]
  
  merged_sce_object$cluster = sce_object_mnn$cluster
  
  # Add in the cluster annotation and cell type classification
  cluster_differentiation_annotation = cluster_annotations[cluster_annotations$Differentiation == d, ]
  cluster_differentiation_annotation = cluster_differentiation_annotation[order(cluster_differentiation_annotation$ClusterID), ]
  cluster_differentiation_annotation$Annotation = make.unique(cluster_differentiation_annotation$Annotation, sep = " ")
  
  merged_sce_object$annotation = sapply(merged_sce_object$cluster, function(i) cluster_differentiation_annotation$Annotation[i])
  merged_sce_object$annotation_category = sapply(merged_sce_object$cluster, function(i) cluster_differentiation_annotation$Annotation.category[i])
  
  # Transfer the reducedDim representation and trim it to the first 2 dim only
  for (reduced_dim in names(reducedDims(sce_object_mnn))) {
    reducedDim(merged_sce_object, reduced_dim) = reducedDim(sce_object_mnn, reduced_dim)[,c(1:2)]
  }
  
  # Fix name for KOLF2 line
  merged_sce_object$vireo_assignment = gsub("KOLF2", "KOLF2.1J", merged_sce_object$vireo_assignment)
  
  merged_sce_object
})
names(differentiations_sce_objects) = differentiations
differentiations_sce_objects
```

## Figures

```{r}
selected_markers = list(
  "cortical" = c("GAP43", "STMN2", "MAP2", "VIM", "VCAN", "FABP7", "PLS3", "METRN", "FGF8", "TTR", "GRIA2", "SLC3A2", "NNAT", "GAD2", "TBR1", "TOP2A", "NEUROG1", "DUSP6"),
  "hypothalamus" = c("GAP43", "STMN2", "VIM", "FABP7", "SLC17A6", "HCRT", "PRRX1", "EMX2", "FGF8", "FOXD1", "GAD2", "SLC32A1", "ZFHX3", "SLC3A2", "TBR1", "CRABP1", "NEUROG1", "WLS", "SLC18A3", "LHX1", "TOP2A", "PCP4", "SST", "POMC"),
  "hNGN2" = c("GAP43", "STMN2", "METRN", "VIM", "SLC17A7", "FABP7", "TOP2A", "ONECUT3", "SLC17A6", "ISL1", "SLC18A3", "GRIA2", "SOX11", "NNAT", "DLL3", "ONECUT2"),
  "hNIL" = c("GAP43", "STMN2", "ONECUT2", "ISL1", "FOXG1", "DLK1", "WLS", "SNCA", "TOP2A", "VIM", "DLL3", "SOX10", "S100B", "PAX6", "DUSP1", "PRRX1", "RAX"))
```

#### Violin plot

```{r, fig.width=6, fig.height=3}
extract_max = function(p){
  ymax = max(ggplot_build(p)$layout$panel_scales_y[[1]]$range$range)
  return(ceiling(ymax))
}

horizontal_violins = lapply(names(differentiations_sce_objects), function(diff) {
  print(diff)
  sce_object = differentiations_sce_objects[[diff]]
  markers = unique(selected_markers[[diff]])
  
  marker_plots = lapply(1:length(unique(selected_markers[["hypothalamus"]])), function(i){
    if (i <= length(markers)) {
      marker = markers[[i]]
      p = plotExpression(sce_object, marker, x="cluster", colour_by = "cluster", point_size=0.1) + 
        theme_nothing() +
        scale_colour_manual(values=color_palette[[diff]]) +
        ylab(marker) + xlab("") +
        theme(legend.position = "none", 
          strip.background = element_blank(),
          strip.text.y = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(), 
          axis.title.x = element_text(size = rel(1), angle=45, vjust = 0.5, hjust=1, margin = margin(t=-3)), 
          axis.text.x = element_text(size = rel(1)), 
          axis.line = element_line(),
          plot.margin = unit(c(-3, 0, -3, 0), "cm")
        ) + coord_flip() + geom_quasirandom_rast(mapping = aes(col=colour_by), alpha=1, size=0.1, width=0.4, groupOnX=TRUE, bandwidth=1, dpi=600, dev = "ragg_png")
      p %>% remove_geom('point', 1)
    } else {
      patchwork::plot_spacer() + theme_nothing() + theme(plot.margin = unit(c(-3, 0, -3, 0), "cm"))
    }
  })
  
  
  marker_plots[[1]] = marker_plots[[1]] +
    theme(axis.text.y=element_text(), axis.ticks.y = element_line(), axis.title.x = element_text()) + xlab("cluster")
  
  xmaxs = purrr::map_dbl(marker_plots, extract_max)
  marker_plots = purrr::map2(marker_plots, xmaxs, function(x,y) x + 
                            scale_y_continuous(breaks = c(y)) + 
                            expand_limits(y = y))
  
  patchwork::wrap_plots(marker_plots, nrow=1)
})
names(horizontal_violins) = names(differentiations_sce_objects)
```

## Suplementary figure

```{r, fig.width=12, fig.height=13}
plot_grid(plotlist = lapply(differentiations, function(d){
  sce_object = differentiations_sce_objects[[d]]
  diff_annotation = cluster_annotations[cluster_annotations$Differentiation == d, ]
  diff_annotation$Annotation = make.unique(diff_annotation$Annotation, sep = " ")
  sce_object$annotation = factor(stringr::str_wrap(paste0(sce_object$cluster, " - ", sce_object$annotation), width=50), levels=stringr::str_wrap(paste0(diff_annotation$ClusterID, " - ", diff_annotation$Annotation), width=50))
  plots = plot_grid(
      remove_geom(plotUMAP(sce_object, colour_by="annotation") + scale_colour_manual(name="Annotation", values = color_palette[[d]]) + ggtitle(paste0(d)) + theme(legend.position = "none") + rasterise(geom_point(mapping = aes(col=colour_by), alpha=0.6, size=0.75), dpi=600), "point", 1),
      get_legend(plotUMAP(sce_object, colour_by="annotation", point_alpha=1) + scale_colour_manual(name="Annotation", values = color_palette[[d]]) + guides(colour = guide_legend(override.aes = list(size=3))) + theme(legend.text = element_text(size=15), legend.title = element_text(size=17), legend.key.size = unit(0.5,"cm"), legend.margin = margin(l=35))),
      horizontal_violins[[d]],
      ncol=3, rel_widths = c(1,1,2)
    )
  
  if (d != differentiations[[length(differentiations)]]) {
    plots = plot_grid(plots, NULL, ncol=1, rel_heights = c(1, 0.1))
  }
  
  plots
}), ncol=1)
ggsave("sup_figure_5.pdf", height = 24, width = 24, units = "in", colormodel = "cmyk")
ggsave("sup_figure_5.png", height = 24, width = 24, units = "in", dpi = 600)
```

