---
title: "hiPSCi cell line evaluation (iPSC) - 05 Feature selection"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=4, fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)
```

```{r, warning=FALSE}
# Load required packages
library(DropletUtils)
library(Matrix)
library(scater) 
library(scran)
library(BiocSingular)
library(BiocParallel)
library(pheatmap)
library(plotly)
library(batchelor)
library(viridis)
library(reshape2)
library(SingleCellExperiment)
library(cowplot)
library(dplyr)
theme_set(theme_cowplot())
ncores = 3
mcparam = MulticoreParam(workers = ncores)
register(mcparam)
```

For this experiment, we have 1 sample composed of all 8 iPSC cell line (undifferentiated) which have been multiplexed by genotype. 

# Feature selection

Since we only have one iPSC samples, we do not need to do any batch correction, but we will still need to perform feature selection for dimensionality reduction and clustering.

```{r}
merged_sce_object = readRDS("dc_cell.rds")

# To properly perform batch correction, we need to remove doublets first
merged_sce_object = merged_sce_object[, colData(merged_sce_object)$final_doublet_call == "Singlet"]
```

For the modeling of gene-variance relationship, we will again remove MT and Ribosomal genes due to the issue we see during doublet detection.

```{r, fig.height=3}
set.seed(1000)
non_mito_ribo_genes = rownames(merged_sce_object)[grep(pattern = "^(MT-|RP[LS]|MALAT1)", rownames(merged_sce_object), invert = TRUE)]
dec_combined = modelGeneVar(merged_sce_object, subset.row=non_mito_ribo_genes, block=merged_sce_object$Sample)
hvg_combined = getTopHVGs(dec_combined)
length(hvg_combined)
```

```{r}
fit = metadata(dec_combined)
ggplot(data = data.frame(mean = fit$mean, var = fit$var, name=ifelse(fit$mean > 4 | fit$var > 2, names(fit$mean), NA)),
                mapping = aes(x = mean, y = var, label=name)) +
  geom_point() +
  labs(x = "Mean log-expression", y = "Variance of log-expression") +
  stat_function(fun=fit$trend, colour="dodgerblue", size=1)
```

Let's now calculate the UMAP for the single dataset.

```{r}
set.seed(1000)
merged_sce_object = runPCA(merged_sce_object, subset_row=hvg_combined)
merged_sce_object = runUMAP(merged_sce_object, ncomponents=2, dimred="PCA", spread = 1, min_dist = 0.4)
```

```{r, fig.width=5, fig.height=5}
plotUMAP(merged_sce_object, colour_by="Sample") + ggtitle("Original")
```

```{r}
saveRDS(merged_sce_object, "all_corrections.rds")
```


