---
title: "hiPSCi cell line evaluation (hypothalamic) - 02 Quality Control Analysis"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)
```

```{r, warning=FALSE}
# Load required packages
library(DropletUtils)
library(Matrix)
library(scater) 
library(scran)
library(BiocSingular)
library(BiocParallel)
library(pheatmap)
library(plotly)
library(batchelor)
library(viridis)
library(reshape2)
library(SingleCellExperiment)
library(cowplot)
library(dplyr)
ncores = 3
mcparam = MulticoreParam(workers = ncores)
register(mcparam)
```

For this experiment, we have 2 samples which have been multiplexed by genotype. Unlike the cortical and dopaminergic differentiation where all the cell lines were pooled together before differentiation, in hypothalamic differentiation we performed differentiation for each cell line individually.

We also do not have any user replicate in hypothalamic differentiation and only have technical replicate. As such the set up of the sample libraries are as follow:

- Sample 1: Hypothalamic cell line 1-9 A
- Sample 2: Hypothalamic cell line 1-9 B

# Pre-processsing

```{r}
sce_object = readRDS("called_cell.rds")
```

## Vireo demultiplexing

```{r}
vireo_assignments = lapply(paste0(list.files(pattern="^Sample_.*vireo_vartrix_genotype"), "/donor_ids.tsv"), read.table, header=TRUE, row.names = 1, stringsAsFactors = FALSE)
lapply(vireo_assignments, head)
```

```{r}
vireo_assignments_called_only = lapply(1:length(sce_object), function(i) {
  vireo_assignments[[i]][colData(sce_object[[i]])$Barcode,]
})
lapply(vireo_assignments_called_only, function(x) table(x$donor_id))
```

```{r}
sce_object = lapply(1:length(sce_object), function(i) {
  sce = sce_object[[i]]
  current_donor_id = vireo_assignments_called_only[[i]]$donor_id
  current_donor_id[current_donor_id == "donor7"] = "HUES9"
  colData(sce)$vireo_assignment = current_donor_id
  sce
})

sce_object
```

```{r, fig.width=2.5, fig.height=4}
cell_lines = c("KOLF2-ARID2-A02", "KUCG3-C1", "LNGPI1-C1","NCRM1-C6","NCRM5-C5","NN0003932-C3","PGP1-C2", "HUES9","doublet", "unassigned")
vireo_count_df = Reduce(rbind, lapply(sce_object, function(x) data.frame(table(colData(x)$vireo_assignment))))
vireo_count_df$Var1 = factor(vireo_count_df$Var1, levels = cell_lines)
colnames(vireo_count_df) = c("Cell line", "Count")
vireo_count_df$Sample = Reduce(c, lapply(1:length(sce_object), function(i) rep(i, length(cell_lines))))

theme_set(theme_classic())
ggplot(data=vireo_count_df, aes(x=Sample, y=Count, fill=`Cell line`)) + 
  geom_bar(stat="identity") + 
  scale_x_discrete() +
  labs(x=NULL)+
  ggtitle("Cell line assignment") +
  geom_text(aes(label=Count), position=position_stack(vjust=0.5), colour="white") +
  labs(fill="")
```

## Merging all samples

Let's now merge all the samples into one SCE object.

```{r}
merged_sce_object = Reduce(cbind, sce_object)
colData(merged_sce_object)$Sample = gsub("/.+", "", colData(merged_sce_object)$Sample)
merged_sce_object
```

## Gene annotation

We want to now annotate the gene with their location so we can determine the mitochondrial genes. These MT genes will be used in QC as a measure of cell viability.

```{r}
library(EnsDb.Hsapiens.v86)
GRCh38_location = mapIds(EnsDb.Hsapiens.v86, keys=rownames(rowData(merged_sce_object)), column="SEQNAME", keytype="GENEID")
rowData(merged_sce_object)$Chromosome = GRCh38_location
rowData(merged_sce_object)$Symbol = gsub("GRCh38_", "", rowData(merged_sce_object)$Symbol)

rownames(merged_sce_object) = uniquifyFeatureNames(rowData(merged_sce_object)$ID, rowData(merged_sce_object)$Symbol)
rowData(merged_sce_object)
```

#  Quality control

We will use the following metrics for performing quality control check:

1. Library size
2. Number of expressed features
3. Proportion of mitochondrial reads (in lieu of spike in control)

```{r}
mito_genes = which(rowData(merged_sce_object)$Chromosome=="MT")
ribo_genes = which(grepl("^RP[LS]", rowData(merged_sce_object)$Symbol, ))
names(ribo_genes) = rowData(merged_sce_object)$Symbol[ribo_genes]
merged_sce_object = addPerCellQC(merged_sce_object, subsets = list(Mito=mito_genes, Ribo=ribo_genes))
```

## Library size

Let's look at the library size to see if there is similar distribution across all samples. After downsampling, we expect to see similar distribution across all samples.

```{r, fig.height=4, fig.width=5}
plotColData(merged_sce_object, x="Sample", y="sum", colour_by = "vireo_assignment") +
  scale_y_log10() +
  labs(y = "UMI count")
```

The downsampling seems to work quite nicely. We can now do the QC across batch. We will also want to remove cell that have too high of a UMI count on top of cell that have too low of a UMI count. 

```{r, fig.height=4, fig.width=5}
qc.lib.size = isOutlier(merged_sce_object$sum, log=TRUE, type="both", batch = merged_sce_object$Sample)

plotColData(merged_sce_object, x="Sample", y="sum", colour_by = I(qc.lib.size)) +
  scale_y_log10() +
  labs(y = "UMI count")
```

Now that the cells have been downsampled, the across sample thresholding makes much more sense!. Note that the orange colours are the outlier.

```{r}
plotColData(merged_sce_object, x="Sample", y="sum", colour_by = I(qc.lib.size), other_fields = "vireo_assignment") +
  scale_y_log10() +
  labs(y = "UMI count") + 
  facet_wrap(~vireo_assignment)
```


## Expressed features

Let's now look at the expressed features distribution for each samples to see if we see different distribution.

```{r, fig.height=4, fig.width=5}
plotColData(merged_sce_object, x="Sample", y="detected", colour_by = "vireo_assignment") +
  scale_y_log10() +
  labs(y = "Number of expressed genes")
```

We see that the ranges of the distribution are quite similar.

```{r, fig.height=4, fig.width=5}
qc.expr.genes = isOutlier(merged_sce_object$detected, log=TRUE, type="both", batch = merged_sce_object$Sample)

plotColData(merged_sce_object, x="Sample", y="detected", colour_by = I(qc.expr.genes)) +
  scale_y_log10() +
  labs(y = "Number of expressed genes")
```

Orange = outlier.

```{r}
plotColData(merged_sce_object, x="Sample", y="detected", colour_by = I(qc.expr.genes), other_fields = "vireo_assignment") +
  scale_y_log10() +
  labs(y = "Number of expressed genes") + facet_wrap(~vireo_assignment)
```

## Cell complexity thresholding

Let's combine both the UMI count and number of gene expressed into a single plot to look at the cell complexity thresholding.

```{r, fig.height=8, fig.width=10}
plotColData(merged_sce_object, x="sum", y="detected", colour_by = I(qc.expr.genes | qc.lib.size), other_fields = "Sample") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "UMI count", y = "Number of expressed genes") +
  facet_wrap(~Sample, ncol = 2)
```

## Mitochondrial gene expression

Let's look at per sample mitochondrial gene expression

```{r}
plotColData(merged_sce_object, x="Sample", y="subsets_Mito_percent", colour_by = "vireo_assignment", other_fields = "Sample")
```

For mito percentage, it seems that QC across sample will likely not turn out well given the still very different distribution between the samples. So we will do the QC per sample here.

```{r}
qc.high.mito = isOutlier(merged_sce_object$subsets_Mito_percent, nmads=3, type="higher", batch = merged_sce_object$Sample)

plotColData(merged_sce_object, x="Sample", y="subsets_Mito_percent", colour_by = I(qc.high.mito), other_fields = "Sample") +
  labs(y = "Mitochondrial read percentage")
```

```{r}
plotColData(merged_sce_object, x="Sample", y="subsets_Mito_percent", colour_by = I(qc.high.mito), other_fields = "vireo_assignment") +
  labs(y = "Mitochondrial read percentage") + facet_wrap(~vireo_assignment)
```

```{r, fig.width=10, fig.height=8}
plotColData(merged_sce_object, x="sum", y="subsets_Mito_percent", colour_by = I(qc.high.mito), other_fields = "Sample") +
  labs(x="UMI count", y = "Mitochondrial read percentage") +
  facet_wrap(~Sample, ncol=2)
```

## Ribosomal gene expression

```{r}
plotColData(merged_sce_object, x="Sample", y="subsets_Ribo_percent", colour_by = "vireo_assignment", other_fields = "Sample")
```

```{r}
plotColData(merged_sce_object, x="Sample", y="subsets_Ribo_percent", colour_by = "vireo_assignment", other_fields = "vireo_assignment") + facet_wrap(~vireo_assignment)
```

For the hypothalamus differentiation, we can clearly see a presence of double hump in 3 cell lines - HUES9, LNGP1 and PGP1 - and some presence of this second hump in other cell lines as well.

```{r}
plotColData(merged_sce_object, x="Sample", y="subsets_Ribo_percent", colour_by = I(log(merged_sce_object$sum)), other_fields = "vireo_assignment") + facet_wrap(~vireo_assignment)
```

We are more likely to see cells with low UMI count in the lower hump, but overall, there is no trend to suggest that high UMI count results in higher ribosomal content.

```{r}
qc.high.ribo = isOutlier(merged_sce_object$subsets_Ribo_percent, nmads=3, type="higher", batch = merged_sce_object$Sample)

plotColData(merged_sce_object, x="Sample", y="subsets_Ribo_percent", colour_by = I(qc.high.ribo), other_fields = "Sample") +
  labs(y = "Ribosomal read percentage")
```

We see a small amount of cell identified as outlier by ribosomal content.

# Outlier determination

Since we see issue with mitochondrial and mitochondrial genes having high average expression and high variance, let's see if the approach we used with removing cell with too many UMI (and with too many genes, though there is only a few here) have effect on cells that get filtered.

```{r, fig.height=8, fig.width=10}
plotColData(merged_sce_object, x="sum", y="subsets_Mito_percent", colour_by = I(qc.high.mito | qc.expr.genes | qc.lib.size), other_fields = "Sample") +
  labs(x="UMI count", y = "Mitochondrial read percentage") +
  facet_wrap(~Sample, ncol=2)
```
We will use some qc metrics (UMI count, detected genes, MT content) for determining which cell to keep.

```{r}
discard_reasons = data.frame(sample=merged_sce_object$Sample, cell_line=merged_sce_object$vireo_assignment, outlier_umi_count=qc.lib.size, outlier_expressed_genes=qc.expr.genes, high_mito_content=qc.high.mito, discarded=qc.high.mito | qc.expr.genes | qc.lib.size)

group_by(discard_reasons, sample) %>% summarise(outlier_umi_count = sum(outlier_umi_count), outlier_expressed_genes = sum(outlier_expressed_genes), high_mito_content = sum(high_mito_content), discarded = sum(discarded), total_cells = n())
```

Let's have a quick look at the per cell line discard statistics.

```{r}
group_by(discard_reasons, sample, cell_line) %>% summarise(outlier_umi_count = sum(outlier_umi_count), outlier_expressed_genes = sum(outlier_expressed_genes), high_mito_content = sum(high_mito_content), discarded = sum(discarded), total_cells = n())
```

```{r}
merged_sce_object = merged_sce_object[, !(qc.high.mito | qc.expr.genes | qc.lib.size | merged_sce_object$vireo_assignment == "unassigned")]
merged_sce_object
```

```{r}
saveRDS(discard_reasons, "discard_reasons.rds")
saveRDS(merged_sce_object, "qc_cell.rds")
```

